---
title: Marascuillo Procedure
author: 
date: '2017-11-22'
slug: Marascuillo Procedure
categories:
  - Marascuillo Procedure
tags: [R, Marascuillo]
---

## Preface
In an attempting to determine which variables are explanatory or response for a data set of all categorical variables, I came across the marascuilo procedure, which isn't used per se for determining whether a variable is explanatory or response.  This data are from the General Social Survey.     

```{r set-options, echo=FALSE, cache=FALSE}
options(width = 10000)
options(warn=1)
```

```{r load, echo=FALSE}
load("/Users/michalondras/education/coursera/inferential-statistics/final\ project/gss.Rdata")
```

```{r decades, echo=FALSE}
# creates new grouping variable decade 
gss$decade = ifelse(gss$year<1980, '70s', 
                    ifelse(gss$year<1990 & gss$year>1979, '80s', 
                           ifelse(gss$year<2000 & gss$year>1989, '90s', 
                                  ifelse(gss$year<2010 & gss$year>1999, 
                                         '00s', '10s'))))
```

```{r contigency, echo=FALSE}
source("http://pcwww.liv.ac.uk/~william/R/crosstab.r")

# creates contigency tables for decades 
contigency_table = function(temp){
  return(crosstab(temp, row.vars = "degree", 
                     col.vars = c("sex", "natenvir"), 
                     addmargins = F))
}

decade = function(temp, ct){
  decades = c("70s", "80s", "90s", "00s", "10s")      # decades of interest
  return(lapply(decades, function(index) ct(temp[temp$decade == index, ])))
}
  
decades = decade(gss, contigency_table)
```

## Categorical Data of Interest
The categorical variables of interest are:

+ Nominal (`sex`)
    + Male or Female
+ Ordinal (`degree`)
    + Little High School, High School, Junior College, Bachelors, and Graduate
+ Ordinal (`natenvir`)
    + Too Little, About Right, and Too Much

Statisticians prefer to express the intent of their experiments, studies before performing a set of analyses.  Unfortunately, a lot of the time a statistician is asked to make sense of data after the fact.  Though not a statistician but an applied mathematician, I am attempting to make sense of data after the fact by asking myself some questions. 

**Which of these variables could be explanatory or response variables?**

Sex should be an explanatory variable. It should be pretty consistent among the decades in the US, as sex-selective abortion is not wide-scaled practiced.  

A contingency table of the proportions of males and females surveyed per decade is displayed below. 

```{r epa}
crosstab(gss, row.vars = "decade", col.vars = "sex", type='r')
```
**Are these proportions changing with time?** 

The Marascuillo procedure can be used to compare pairwise differences between the decades.  A pair proportion difference is statistically significant if its value exceeds the critical range value. Here, we state that the following hypotheses:

+ $H_o$: there is no difference between the pairwise proportion differences
+ $H_a$: there is a difference between the pairwise proportion differences

The critical range will be $\chi^2$ statistic of 9.49 (0.95 CI, degrees of freedom = 4). 

```{r marascuillo}
gender = crosstab(gss, row.vars = "decade", col.vars = "sex", type='f')

marascuillo_procedure = function(p){
  n = length(p[, c(1)]) - 1
  prop = p[,1][1:n]/p[,2][1:n]
  names = row.names(p)
  N = length(prop)
  value = critical.range = compare= c()
  
  ## Compute critical values; small data set; efficiency non-issue
  for (i in 1:(N - 1))
     { for (j in (i + 1):N){
       value = c(value,(abs(prop[i] - prop[j])))
       critical.range = c(critical.range, 
          sqrt(qchisq(.95,4))*sqrt(prop[i]*(1-prop[i])/p[i,2] + prop[j]*(1-prop[j])/p[j,2]))
        compare = c(compare, paste(names[i], names[j], sep = '-'))
      }
  }
  
  return(data.frame(values = round(value, 3),
                    critical = round(critical.range,3), 
                    comparison = compare))
}  

marascuillo_procedure(gender$table[, c(1,3)])
```
The statistically significant differences are these pairwise comparisons: 80s and 00s; 70s and 80s; and 70s and 90s.  The decade demarcation could afford some tweaking.  Looking at year-by-year proportion differences, I observed that the years 1972, 73, and 84, when compared to other years, are statistically different. 

```{r year-by-year}
# list of year contigency tables 
gender.yr = crosstab(gss, row.vars = "year", col.vars = "sex", type='f')
gender.prop.yr = marascuillo_procedure(gender.yr$table[, c(1,3)])
gender.prop.yr[gender.prop.yr$values > gender.prop.yr$critical, ]
```
**What if I were to elminate them?** 

if I eliminate those years,  then I see that there are no pairwise decade-by-decade differences in proportions that are significant.  The years of potential exclusion were years during which additional tests, surveys were done that focused on increasing information from demographics undocumented in the past.    

```{r decade-by-decade}
gender = crosstab(gss[gss$year > 1973 & gss$year != 1984,], 
                  row.vars = "decade", col.vars = "sex", type='f')
marascuillo_procedure(gender$table[, c(1,3)])
```

Gender could be an explanatory variable.  More of an assessment is needed to determine whether exclusion of those years is warranted and perform subsequent analyses. 

## References: 
1. NIST: _Comparing multiple proportions: The Marascuillo procedure_,  
http://www.itl.nist.gov/div898/handbook/prc/section4/prc474.htm
