---
title: Marascuillo Procedure
author: 
date: '2017-11-22'
slug: Marascuillo Procedure
categories:
  - R
tags: []
---

## Setup
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 10000)
options(warn=1)
```

### Load packages
```{r load-packages, message=FALSE}
```

### Load data
```{r load, echo=FALSE}
setwd("/Users/michalondras/education/coursera/inferential-statistics/final\ project/")
load("gss.Rdata")
```

```{r decades}
# creates new grouping variable decade 
gss$decade = ifelse(gss$year<1980, '70s', 
                    ifelse(gss$year<1990 & gss$year>1979, '80s', 
                           ifelse(gss$year<2000 & gss$year>1989, '90s', 
                                  ifelse(gss$year<2010 & gss$year>1999, 
                                         '00s', '10s'))))
```

```{r contigency}
source("http://pcwww.liv.ac.uk/~william/R/crosstab.r")

# creates contigency tables for decades 
contigency_table = function(temp){
  temp_table = crosstab(temp, row.vars = "degree", 
                     col.vars = c("sex", "natenvir"), 
                     addmargins = F)
  return(temp_table)
}

decade = function(temp, ct){
  # decades of interest
  decades = c("70s", "80s", "90s", "00s", "10s")
  
  # list of contigency tables 
  data.list = list()
  for (index in decades){
    data.list[[index]] = ct(temp[temp$decade == index, ])
  }
  return(data.list)
}
  
decades = decade(gss, contigency_table)
```

#### Categorical Data of Interest
The categorical variables of interest are:

+ Nominal (`sex`)
    + Male or Female
+ Ordinal (`degree`)
    + Little High School, High School, Junior College, Bachelors, and Graduate
+ Ordinal (`natenvir`)
    + Too Little, About Right, and Too Much

**Which of these variables are explanatory or response variables?** Sex should be an explanatory variable. It should be pretty consistent among the decades.  A contigency table of the proportions of males and females surveyed per decade is displayed below. 
```{r epa}
crosstab(gss, row.vars = "decade", col.vars = "sex", type='r')
```
**Are these proportions changing with time?** The Marascuillo procedure can be used to compare pairwise differences between the decades.  A pair proportion difference is statistically significant if its value exceeds the critical range value. Here, we state that the following hypotheses:

+ $H_o$: there is no difference between the pairwise proportion differences
+ $H_a$: there is a difference between the pairwise proportion differences

The critical range will be $\chi^2$ statistic of 9.49 (0.95 CI, degrees of freedom = 4). 

```{r marascuillo}
# TODO shiny data.table with highlighted values
gender = crosstab(gss, row.vars = "decade", col.vars = "sex", type='f')

marascuillo_procedure = function(p){
  n = length(p[, c(1)]) - 1
  prop = p[,1][1:n]/p[,2][1:n]
  names = row.names(p)
  N = length(prop)
  value = critical.range = compare= c()
  
  ## Compute critical values.
  for (i in 1:(N-1))
     { for (j in (i+1):N)
      {
       value = c(value,(abs(prop[i]-prop[j])))
       critical.range = c(critical.range,
        sqrt(qchisq(.95,4))*sqrt(prop[i]*(1-prop[i])/p[i,2] + prop[j]*(1-prop[j])/p[j,2]))
        compare = c(compare, paste(names[i], names[j], sep = '-'))
      }
     }
  return(data.frame(values = round(value, 3),
                    critical = round(critical.range,3), 
                    comparison = compare))
}  

marascuillo_procedure(gender$table[, c(1,3)])
```
The statistically significant differences are these pairwise comparisions: 80s and 00s; 70s and 80s; and 70s and 90s.  My decade demarcation will need a little tweaking.  Looking at year-by-year proportion differences, I see that the years 1972, 73, and 84, when compared to other years, are statistically different. 

```{r year-by-year}
# list of year contigency tables 
gender.yr = crosstab(gss, row.vars = "year", col.vars = "sex", type='f')
gender.prop.yr = marascuillo_procedure(gender.yr$table[, c(1,3)])
gender.prop.yr[gender.prop.yr$values > gender.prop.yr$critical, ]
```
**What if I were to elminate them?** Eliminating those years,  I see that there are no pairwise decade-by-decade differences in proportions that are significant.  

```{r decade-by-decade}
gender = crosstab(gss[gss$year > 1973 & gss$year != 1984,], 
                  row.vars = "decade", col.vars = "sex", type='f')
marascuillo_procedure(gender$table[, c(1,3)])
```

Gender could be an explanatory variable. 

1. NIST: _Comparing multiple proportions: The Marascuillo procedure_, 
http://www.itl.nist.gov/div898/handbook/prc/section4/prc474.htm


